---
title: "module-09-notes"
output: html_document
date: "2024-01-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Load "Country Data 2016" data set in from a URL and create a data frame 
```{r}
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/Country-Data-2016.csv"
d <- read_csv(f, col_names = TRUE)
names(d)
```
## Attach function 
Using the attach function makes it so you don't have to use the $ to call on a variable from a dataset 
Tony doesn't recommend using this because if you're working with 2 datasets with variables of the same name, R can get confused 

## Tidy data 
A table is "tidy" when each variable is in its own column, and each observation is in its own row 

### How many variables and how many observations?
```{r}
str(d)
```

## How many countries?
```{r}
str(d$country)
```
## Median area and population size for all countries in dataset 
```{r}
median(d$area, na.rm = TRUE)
median(d$population, na.rm = TRUE)
```
## Create a new density variable in your data frame which is population/area
```{r}
density <- d$population/d$area
d <- cbind(d, density)
d
```
Another way to do the same thing: d$density <- d$population/d$area
Or d <- d |> mutate(density = population/area)

## What are the 10 most dense countries? 
```{r}
order(d$density, decreasing = TRUE)
```
```{r}
d$country[c(130, 145, 97, 99, 84, 17, 135, 137, 24, 18)]
```
## What are the 10 least dense countries?
```{r}
order(d$density, decreasing = FALSE)
```
```{r}
d$country[c(206, 86, 70, 175, 146, 245, 76, 152, 13, 101)]
```

