---
title: "Feb 7 Notes"
output: html_document
date: "2024-02-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
```

# Class 7: Flow Control, Looping, Functions and the Wordle Coding Challenge 
By Monday - read thru and work examples in Module 11, work on finding paper and data set for data analysis replication assignment, submit Exercise 3, start looking at Exercise 4

```{r}
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/IMDB-movies.csv"
d <- read_csv(f, col_names = TRUE)
view(d)
```

## Conditionals 

if_else(test, do this if true, do this if false) 

case_when(
  case1 ~ result1,
  case2 ~ result2,
  case3 ~ result3,
  TRUE ~ all others
)

### Add new column to the data frame that includes all movies where one of the genres listed is "comedy," then count the total number of comedy movies 
```{r}
d <- d %>% 
  mutate(comedy = if_else(str_detect(genres, "Comedy"), TRUE, FALSE)) 
  
s <- d %>% summarise(count = n(), num_comedy = sum(comedy, na.rm = TRUE))
```

### Add new variable for "ranking," where ranking = "low" if averageRating = 0 to 3.3, "med" if averageRating > 3.3 to 6.7, and high if averageRangking > 6.7 to 10 
```{r}
d <- d %>%
  mutate(ranking = case_when(
    averageRating <= 3.3 ~ "Low",
    averageRating <= 6.7 ~ "Med",
    averageRating > 6.7 ~ "High"
  ))
d
```

### Show average run time for movies per ranking category 
```{r}
s2 <- d %>% 
  group_by(ranking) %>% 
  summarise(count = n(), avg_time = mean(runtimeMinutes, na.rm = TRUE))
s2
```

## Loops
for (i in 1:...) {do this} 

### Example loop: Prints numbers 1 thru 28938 
```{r}
for (i in 1:nrow(d)){
  print(i)
}
```

### Loop thru all the rows in the data set and print out the cumulative running time of the movies
```{r}
t <- 0
for (i in 1:nrow(d)){
  t <- ifelse(is.na(d[i,c("runtimeMinutes")]), t, t + d[i,c("runtimeMinutes")])
  print(t)
}
```
### Easier way to get the total:
```{r}
cumsum(replace_na(d$runtimeMinutes, 0))
```
