---
title: "Module 10 Notes"
output: html_document
date: "2024-02-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages("tidyverse")
```
```{r}
library(tidyverse)
```
```{r}
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/KamilarAndCooperData.csv"
d <- read_csv(f, col_names = TRUE)  # creates a 'tibble'
head(d)
```
## Creating boxplots using base R and superimpose individual data points onto plot 
"jitter" ensures there is room between each plot point so they're not all in a vertical line
add superimposes the points
```{r}
boxplot(log(d$Body_mass_female_mean))
stripchart(log(d$Body_mass_female_mean),
           method = "jitter",
           col = "blue",
           vertical = TRUE,
           add = TRUE)
```

```{r}
boxplot(data = d, log(Body_mass_female_mean) ~ Family)
stripchart(data = d, log(Body_mass_female_mean) ~ Family,
           method = "jitter",
           col = "green",
           vertical = TRUE,
           add = TRUE)
```
## Using ggplot2 - grammar of graphics

ggplot(data = ..., mapping = aes(...)) + 
  geom ...(...) + 
  theme(...) + ... 
  
```{r}
p <- ggplot(data = d, aes(x = "", y = log(Body_mass_female_mean))) + geom_boxplot(na.rm = TRUE) # x is a dummy variable here 
p
```
```{r}
p <- p + geom_jitter(color = "blue", width = 0.1)
p
```

Replacing the dummy variable x with Family variable from data set 
axis.text.x = element_text(angle = 90) rotates the x axis labels 90 degrees for readability 
ylab changes the y axis label text
```{r}
p <- ggplot(data = d, aes(x = Family, y = log(Body_mass_female_mean))) + geom_boxplot(na.rm = TRUE) + geom_jitter(color = "blue", width = 0.1)
p <- p + theme(axis.text.x = element_text(angle = 90))
p <- p + ylab("log(Female Body Mass)")
p
```
## Creating histograms in base R 
```{r}
hist(log(d$Body_mass_female_mean)) #freq = TRUE, y axis is a count 
hist(log(d$Body_mass_female_mean), freq = FALSE) #freq = FALSE, y axis equals density (height of each bar equals 1 total)
lines(
  density(log(d$Body_mass_female_mean), na.rm = TRUE),
  col = "blue") # adds density function on top of histogram 
abline(v = quantile(
       log(d$Body_mass_female_mean),
       prob = c(0.05, 0.95),
       na.rm = TRUE
       ),
col = "red") #adds red lines at 5% and 95% confidence intervals 
```

##Creating histograms with ggplot2
```{r}
p <- ggplot(data = d,
            aes(log(Body_mass_female_mean))
            )
p <- p + geom_histogram(binwidth = 2)
(p + geom_histogram(bins = 9)) #wrapping code in parantheses prints plot without reassigning variable p 
p <- (p + geom_histogram(bins = 9, aes(y = ..density..)) + geom_density()) #shows aesthetics of histogram as a density curve
p <- p + geom_vline(xintercept = quantile(
  log(d$Body_mass_female_mean),
  prob = c(0.025, 0.975),
  na.rm = TRUE
), 
color = "red"
)
```

## Scatterplots
```{r}
plot(x = log(d$Body_mass_female_mean),
     y = log(d$Brain_Size_Female_Mean)
     )
```
```{r}
n <- ggplot(data = d, aes(x = log(Body_mass_female_mean), y = log(Brain_Size_Female_Mean)))
n <- n + geom_point(na.rm = TRUE)
n <- n + geom_point(aes(color = factor(Family)), na.rm = TRUE)
n
```
### Modify axes
```{r}
n <- n + xlab("log(Female Body Mass)") + ylab("log(Female Brain Size") + theme(legend.position = "bottom", legend.title = element_blank())
n <- n + geom_smooth(method = "lm", na.rm = TRUE) #add linear regression line with standard error ribbon to plot 
n
```
## Data wrangling 
Filtering using base R
```{r}
s <- d[d$Family == "Hominidae" & d$Mass_Dimorphism > 2,] #applying joint criteria to data set and filtering out lines based on that 
s
```

Filtering using {dplyr}
```{r}
s <- filter(d, Family == "Hominidae" & Mass_Dimorphism >2)
s
```

Selecting... pulling out columns of data set using {dplyr}
```{r}
s <- select(d, Family, Genus, Body_mass_female_mean)
s
```

Sorting using {dplyr}
```{r}
s <- arrange(d, Family, Genus, desc(Body_mass_male_mean))
s
```

Summarizing by group using {dplyr}... each line of code does the same thing 
```{r}
s <- summarise(
  group_by(d, Family),
  avgF = mean(Body_mass_female_mean, na.rm = TRUE)
)

s  <- group_by(d, Family) %>%
  summarise(avgF = mean(Body_mass_female_mean), na.rm = TRUE)

s <- d %>%
  group_by(Family) %>%
  summarise(avgF = mean(Body_mass_female_mean, na.rm = TRUE))
s
```

Mutate is used to create a new variable 
```{r}
s <- s %>% mutate(new = paste0(Family, "-", avgF))
s
```

