---
title: "Intro to Spatial Data"
author: "Claire Brandes"
date: "April 17, 2024"
output: 
  html_document: 
    highlight: tango
---

**Loading required packages:**

```{r message=FALSE}
easypackages::libraries("tidyverse", 
                        "sp", 
                        "sf", 
                        "cowplot", 
                        "adehabitatHR")
```

**Loading data set:**

```{r message=FALSE}
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/spider_monkeys.csv"
d <- read_csv(f, col_names = TRUE)
```

## Projections and Coordinate Reference Systems

**Converting data frame to spatial data object:**

```{r}
# Project and plot focal points
crs_original <- "+proj=longlat +datum=WGS84" 

# x, y values represent the longitude and latitude, respectively and match the names of the fields in our original data table
coordinates(d) <- c("Longitude", "Latitude")  

# Note that the data need to have an initial projection assigned in order to be transformed
# Here we know that our initial coordinates are in latitude-longitude
proj4string(d) <- CRS(crs_original)  
crs_target <- "+proj=utm +zone=18 + south +datum=WGS84 +units=m"
d_sp <- spTransform(d, CRS(crs_target))
d_sf <- st_as_sf(d_sp, CRS("+proj=utm +zone=18 + south +datum=WGS84 +units=m"))
plot(d_sp, pch = 19, main = "Focal Points")
```

**Loading in two other data sets: tree coordinates, and single point location for the Tiputini Biodiversity Station**

```{r message=FALSE}
# Point data for trees through 2012
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/trees_2012.csv"
trees <- read_csv(f, col_names = TRUE)

# Point data for TBS field station
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/TBS_coordinates.csv"
station <- read_csv(f, col_names = TRUE)
```

**Converting data sets to spatial objects:**

```{r}
coordinates(trees) <- c("FINAL_LON", "FINAL_LAT")
proj4string(trees) <- CRS("+proj=longlat +datum=WGS84")
trees_sp <- spTransform(trees, CRS("+proj=utm +zone=18 + south +datum=WGS84 +units=m"))
trees_sf <- st_as_sf(trees_sp, CRS("+proj=utm +zone=18 + south +datum=WGS84 +units=m"))
plot(trees_sp, pch = 19, main = "Feeding Trees")

coordinates(station) <- c("Longitude", "Latitude")
proj4string(station) <- CRS("+proj=longlat +datum=WGS84")
station_sp <- spTransform(station, CRS("+proj=utm +zone=18 + south +datum=WGS84 +units=m"))
station_sf <- st_as_sf(station_sp, CRS("+proj=utm +zone=18 + south +datum=WGS84 +units=m"))
```

**Point location of mineral lick used by primate group:**

```{r message=FALSE}
# Point data for the mineral lick
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/mineral_lick.csv"
mineral_lick <- read_csv(f, col_names = TRUE)

mineral_lick_sf <- st_as_sf(mineral_lick, coords = c("x_proj", "y_proj"), crs = 32718)
mineral_lick_sp <- as(mineral_lick_sf, "Spatial")
```

**Shapefile for Ecuador:**

```{r message=FALSE}
# Load shapefile, convert to sp/sf
f <- "/Users/clairebrandes/Desktop/UT/Applied Data Analysis/Repos/ada-2024-notes/Spatial Data/ec_provinces.shp"
ecuador <- st_read(f)

ecuador_sp <- as(ecuador, "Spatial")
proj4string(ecuador_sp) <- CRS("+proj=utm +zone=18 + south +datum=WGS84 +units=m")
ecuador_sf <- st_as_sf(ecuador_sp, CRS("+proj=utm +zone=18 + south +datum=WGS84 +units=m"))

# Load and project shapefile for the TBS trail system, convert to sp/sf
f <- "/Users/clairebrandes/Desktop/UT/Applied Data Analysis/Repos/ada-2024-notes/Spatial Data/trails.shp"
trails <- st_read(f)

trails_sp <- as(trails, "Spatial")
proj4string(trails_sp) <- CRS("+proj=longlat +datum=WGS84")
trails_sp <- spTransform(trails_sp, CRS("+proj=utm +zone=18 + south +datum=WGS84 +units=m"))
trails_sf <- st_as_sf(trails_sp, CRS("+proj=utm +zone=18 + south +datum=WGS84 +units=m"))

# Load and project shapefile for the Tiputini River, convert to sp/sf
f <- "/Users/clairebrandes/Desktop/UT/Applied Data Analysis/Repos/ada-2024-notes/Spatial Data/rio tiputini.shp"
rio <- st_read(f)

rio_sp <- as(rio, "Spatial")
proj4string(rio_sp) <- CRS("+proj=longlat +datum=WGS84")
rio_sp <- spTransform(rio_sp, CRS("+proj=utm +zone=18 + south +datum=WGS84 +units=m"))
rio_sf <- st_as_sf(rio_sp, CRS("+proj=utm +zone=18 + south +datum=WGS84 +units=m"))
```

## Simple Plots

```{r}
# plot TBS point over Ecuador Shapefile and TBS point over trail and river
# shapefiles using ggplot() and sf objects

p1 <- ggplot() + geom_sf(data = ecuador_sf) + geom_sf(data = rio_sf, color = "blue") +
    geom_sf(data = station_sf, color = "red") + xlab("Longitude") + ylab("Latitude") +
    ggtitle("TBS, Ecuador") + coord_sf(crs = 32718)

p2 <- ggplot() + geom_sf(data = trails_sf) + geom_sf(data = rio_sf, color = "blue") +
    geom_sf(data = station_sf, color = "red") + xlab("Longitude") + ylab("Latitude") +
    ggtitle("TBS trail system \n and Tiputini river") + coord_sf(crs = 32718)

cowplot::plot_grid(p1, p2, ncol = 2)
```

## Generating Home Ranges

**Calculating home range area (MCP method):**

MCP approach *was* a standard method for calculating home range in the 90s 

```{r}
polygon <- mcp(d_sp, percent = 100)  #set to 100% to include all points

plot(polygon, main = "Home Range, MCP method, 100%")
```

**Same plot with ggplot:**

```{r}
ggplot(st_as_sf(polygon)) + geom_sf(alpha = 0.5)
```

Excluding some points that may be outliers and not accurately represent the group’s home range. Simply need to select the percentage of points to include (in this case, 95% of points):

```{r}
polygon95 <- mcp(d_sp, percent = 95)

plot(polygon95, main = "Home Range, MCP method, 95%")
```

**As ggplot:**

```{r}
ggplot(st_as_sf(polygon95)) + geom_sf(alpha = 0.5)
```

**Utilization distribution model:** 

```{r}
# kernelUD method
hr_kernel <- kernelUD(d_sp)

hr_kernel95 <- getverticeshr(hr_kernel, percent = 95)
plot(hr_kernel95, main = "Home Range, kernelUD method, 95%")
```

**Plotted with ggplot, including data points:**

Need to specify aesthetics associated with data points here

```{r}
ggplot(st_as_sf(hr_kernel95)) + geom_sf(alpha = 0.5) + geom_sf(data = st_as_sf(d_sp),
    aes(geometry = geometry), alpha = 0.3)
```

