---
title: "Mixed Effects Modeling"
author: "Claire Brandes"
date: "April 10, 2024"
output: 
  html_document: 
    highlight: tango
---

In a general or generalized linear "mixed effects" models **(GLMM)**, we have a response variable, Y, and observations that fall into different factor categories, each with some set of levels (e.g., "sex," with levels "male" and "female", "individual", with different individual names), and we are interested in the effects of the various factors and factor levels on the response variable. 

We can conceptualize these effects as being either **fixed** or **random**. Fixed factors are those that reflect all levels of interest in our study, while random effects are those that represent nly a sample of the levels of interest. 

```{r message=FALSE}
library(readr)
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/chimpgrooming.csv"
d <- read_csv(f, col_names = TRUE)
```

Suppose we're interested in how reproductive condition and parity influence how much grooming a female receives. 

Mixed effects model deals with situation by adding **subject ID** as a random effect in our model. Doing so allows us to address the non-independence issue by estimating **different sets of parameters** for each level of the factor "subject." We can do two kinds of mixed effects models:

Estimate a different intercept for each subject  
Estimate a different intercept and slope for each subject 

Plot grooming received duration in relation to subject ID

```{r}
library(ggplot2)
(p <- ggplot(data = d, aes(x = subject, y = duration)) +
  geom_boxplot()) +
  xlab("Subject") + ylab("Duration") +
  theme_classic()
```

Plot grooming received in relation to reproductive condition 

```{r}
(p <- ggplot(data = d, aes(x = reprocondition, y = duration)) +
  geom_boxplot()) +
  xlab("Reproductive Conditiont") + ylab("Duration") +
  theme_classic()
```

Plot grooming received in relation to reproductive condition *and* parity

```{r}
(p <- ggplot(data = d, aes(x = reprocondition, y = duration, fill = factor(parity))) +
  geom_boxplot() + 
   xlab("Reproductive Condition") +
   labs(fill = "parity") + 
   theme(axis.text.x = element_text(angle = 90)) +
   theme_classic())
```


Overall we see: lots of individual variation in how much grooming is received, variation in response to reproductive condition, limited variation due to parity, variation in response to this combination of mixed effects 

Plot grooming received into relation to both reproductive condition and subject ID

```{r}
(p <- ggplot(data = d, aes(x = reprocondition, y = duration, fill = factor(subject))) +
    geom_boxplot() + xlab("reproductive condition") + labs(fill = "subject") + theme(axis.text.x = element_text(angle = 90)))
```

## Random Intercept Model (Parallel Slopes)

```{r message=FALSE}
library(lme4)
```

Function lmer() is the mixed model equivalent of lm()

```{r}
m <- lmer(data = d, duration ~ reprocondition + parity + (1 | subject))
summary(m)
coef(m)
```


### Inference using LRTs

```{r}
full <- lmer(data = d, duration ~ reprocondition + parity + (1 | subject), REML = FALSE)

reduced <- lmer(data = d, duration ~ parity + (1 | subject), REML = FALSE)

anova(reduced, full, test = "Chisq")
```

Adding productive condition in significantly increases explanatory power of the model 

we run with REML = FALSE because we will be comparing models with different "fixed" effects

### Random Slopes Model 

We can also allow that relationship between the response variable and predictor variable(s) to **vary from subject to subject**. 

The notation (1 + fixed | random) tells the model to estimate differing baseline levels of the response as well as differing responses to the fixed factor in question 

```{r}
m <- lmer(data = d, duration ~ reprocondition + parity + (1 + reprocondition|subject) + (1 + parity|subject), REML = FALSE)
summary(m)
```

