---
title: "Model Selection"
author: "Claire Brandes"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: leonids
    toc: yes
---

# Model Selection

## Continued from last class

Loading packages

```{r message=FALSE}
library(tidyverse)
```

Loading data set

```{r message=FALSE}
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/AVONETdataset1.csv"
d <- read_csv(f, col_names = TRUE)
```

Winnowing data set

```{r}
d <- d %>% select("Species1",
                  "Family1",
                  "Order1",
                  "Beak.Length_Culmen",
                  "Beak.Width",
                  "Beak.Depth",
                  "Tarsus.Length",
                  "Wing.Length",
                  "Tail.Length",
                  "Mass",
                  "Habitat",
                  "Migration",
                  "Trophic.Level",
                  "Trophic.Niche",
                  "Min.Latitude",
                  "Max.Latitude",
                  "Centroid.Latitude",
                  "Primary.Lifestyle",
                  "Range.Size")
```

Making some new variables

```{r}
d <- d %>% mutate(logMass = log(Mass),
                  logRS = log(Range.Size),
                  logBeak = log(Beak.Length_Culmen),
                  logTarsus = log(Tarsus.Length),
                  Migration = as.factor(Migration))
```

Lets compare...

```{r}
m1 <- lm(data = d, logBeak ~ logRS * Migration) # full model
m2 <- lm(data = d, logBeak ~ logRS + Migration) # model without interaction term
m3 <- lm(data = d, logBeak ~ logRS) # model with one predictor
m4 <- lm(data = d, logBeak ~ Migration) # model with one predictor
m5 <- lm(data = d, logBeak ~ 1) # intercept only model 
```

Once we have fitted a full and nested model, we can carry out partial F test to compare particular models using the anova() function, with the nested (reduced) and full models as arguments. The reduced model is included as the first argument and the full model is included as the second argument.  

```{r}
anova(m2, m1, test = "F")
```

Does a model including logRS and Migration have better explanatory power than a model including just logRS? Yes! 

## Exploring "forward" and "backward" selection

### Forward Selection

```{r}
d_new <- d %>% drop_na(c(logBeak, logTarsus, logRS, Primary.Lifestyle, Trophic.Level, Migration))

lm(data = d_new, logBeak ~ logRS + Migration + Trophic.Level + logTarsus + Primary.Lifestyle)
```

**Forward selection** starts with an intercept-only model and then tests which of the predictor variables best improves the goodness of fit. Then, the model is updated by adding that term and tests which of the remaining predictors would further and best improve the fit. 

The R functions add1() and update(), respectively, perform the series of tests and update your fitted regression model. Setting the "test=" argument to "F" includes the partial F statistic value and its significance. The ".~." part of the "scope=" argument means, basically, "what is already there", while the remainder of the "scope=" argument is the list of additional variables you might add for the fullest possible model.

```{r}
m_null <- lm(data = d_new, logBeak ~ 1) # starting with null model (logBeak as a function of the intercept)

add1(m_null, scope = .~. + logRS + Migration + Trophic.Level + Primary.Lifestyle + logTarsus, test = "F")
```

All have explanatory power based on p-values. Variable with the highest F value is adding the most explanatory power to the model. In this case, it is "Primary.Lifestyle." Now adding this factor into the null model:

```{r}
m1 <- update(m_null, formula = .~. + logTarsus)
summary(m1)
```

Now we're going to add one more variable to the model. Let's see what the variable with the next highest explanatory power is:

```{r}
add1(m1, scope = .~. + logRS + Migration + Trophic.Level + Primary.Lifestyle + logTarsus, test = "F")
```

Updating model 1 by adding in Primary.Lifestyle

```{r}
m2 <- update(m1, formula = .~. + Primary.Lifestyle)
summary(m2)
```

Overall still a significant model, multiple R^2^ is increasing 

Repeating pattern again:

```{r}
add1(m2, scope = .~. + logRS + Migration + Trophic.Level + Primary.Lifestyle + logTarsus, test = "F")
```
Updating model:

```{r}
m3 <- update(m2, formula = .~. + Trophic.Level)
summary(m3)
```

Repeating again:

```{r}
add1(m3, scope = .~. + logRS + Migration + Trophic.Level + Primary.Lifestyle + logTarsus, test = "F")
```

Adding logRS to the model

```{r}
m4 <- update(m3, formula = .~. + logRS)
summary(m4)
```

Repeating again:

```{r}
add1(m4, scope = .~. + logRS + Migration + Trophic.Level + Primary.Lifestyle + logTarsus, test = "F")
```

### Backward Selection

**Backward selection** starts with the fullest model you want to consider and then systematically drops terms that do not contribute to the explanatory value of the model. The R functions for this process are drop1() to inspect the partial F test results and update() to update the model.

```{r}
m_full <- lm(data = d_new, logBeak ~ logRS + Migration + Trophic.Level + Primary.Lifestyle + logTarsus) # starting with full model
```

Dropping from full model with test type set to F 

```{r}
drop1(m_full, test = "F")
```

Update by removing Migration, which has the lowest F value

```{r}
m2 <- update(m_full, .~. - Migration)
summary(m2)
```

Overall model is still significant. Now we're going to drop from m2

```{r}
drop1(m2, test = "F")
```

Dropping logRS from the model

```{r}
m3 <- update(m2, .~. - logRS)
summary(m3)
```

Dropping from m3

```{r}
drop1(m3, test = "F")
```

Removing Trophic.Level from the model

```{r}
m4 <- update(m3, .~. - Trophic.Level)
summary(m4)
```

Note: The example above doesn't demonstrate a step in which adding or removing a variable adds more explanatory power to the model. 

