---
title: "Multiple Regression and ANCOVA"
author: "Claire Brandes"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: leonids
    toc: yes
---

```{r}
library(tidyverse)
library(car)
library(jtools)
library(broom)
```

# Multiple Regression

Loading in data sets:

```{r}
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/zombies.csv"
z <- read_csv(f, col_names = TRUE)

f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/AVONETdataset1.csv"
d <- read_csv(f, col_names = TRUE)
```

Winnow AVONET data set:

```{r}
d <- d %>% select("Species1","Family1","Order1","Beak.Length_Culmen","Beak.Width","Beak.Depth","Tarsus.Length","Wing.Length","Tail.Length","Mass","Habitat","Migration","Trophic.Level","Trophic.Niche","Min.Latitude","Max.Latitude","Centroid.Latitude","Primary.Lifestyle","Range.Size")
```

Linear model of height as a function of weight + age + gender:

```{r}
m <- lm(height ~ weight + age + gender, data = z)
summary(m)
```

Tidy data:

```{r}
m |> tidy()
```

Check VIFs:

```{r}
vif(m)
```

## Challenge

With AVONET dataset, winnow to order "Accipitriformes" and run a multiple regression of log (Range.Size) as a sunction of the variables log(Mass) and Primary.Lifestyle

```{r}
d <- d %>% filter(Order1 == "Accipitriformes")

m1 <- lm(log(Range.Size) ~ log(Mass) + Primary.Lifestyle, data = d) 
summary(m1)
```

Generate prediction and confidence intervals (using zombies data set as example:

```{r}
m <- lm(height ~ weight + age + gender, data = z)
summary(m)
```

What is the estimated mean height, in inches, for a 29 year old male who weighs 132 pounds who has survived the zombie apocalypse, and what is the 95% confidence interval around this estimate of mean height?

Calculating estimate by hand: 

```{r}
(132 * 0.140542) + (29 * 0.662458) + (1 * 1.609671) + 33.309791
```

Generating confidence interval: 

```{r}
(ci <- predict(m, newdata = data.frame(age = 29, gender = "Male", weight = 132), interval = "confidence", level = 0.095))
```

Generating prediction interval:

```{r}
(pi <- predict(m, newdata = data.frame(age = 29, gender = "Male", weight = 132), interval = "prediction", level = 0.095))
```

**{jtools} package:**

**effect_plot()** can be used to plot the relationship between (one of) the predictor variables and the response variable while holding the others constant

**plot_sums()** can be used to visualize coefficient estimates and CI values around those terms, including visualizing multiple models on the same plot

```{r}
library(broom.mixed)

effect_plot(m, pred = weight, interval = TRUE, int.type = "confidence", int.width = 0.95, plot.points = TRUE)

plot_summs(m)

plot_summs(m, plot.distributions = TRUE, rescale.distributions = TRUE)
```

# Model Selection

Installing packages and loading/winnowing data set:

```{r}
library("MASS")
library("AICcmodavg")
library("MuMIn")

f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/AVONETdataset1.csv"
d <- read_csv(f, col_names = TRUE)
d <- d %>% select("Species1","Family1","Order1","Beak.Length_Culmen","Beak.Width","Beak.Depth","Tarsus.Length","Wing.Length","Tail.Length","Mass","Habitat","Migration","Trophic.Level","Trophic.Niche","Min.Latitude","Max.Latitude","Centroid.Latitude","Primary.Lifestyle","Range.Size")
```

Make some new variables:

```{r}
d <- d %>% mutate(logMass = log(Mass),
                  logRS = log(Range.Size),
                  logBeak = log(Beak.Length_Culmen),
                  logTarsus = log(Tarsus.Length),
                  Migration = as.factor(Migration))
```

Model evaluation using partial F tests

Let's compare...

```{r}
m1 <- lm(data = d, logBeak ~ logRS * Migration) # full model
m2 <- lm(data = d, logBeak ~ logRS + Migration) # model without interaction term
m3 <- lm(data = d, logBeak ~ logRS) # model with one predictor
m4 <- lm(data = d, logBeak ~ Migration) # model with one predictor
m5 <- lm(data = d, logBeak ~ 1) # intercept only model 

anova(m2, m1, test = "F") # tells us that adding the interaction term does add more explanatory power (based on p-value of 0.004)
```


```{r}
anova(m2, m3, test = "F") # before next time - explore filtering data frame to not be missing data passed to linear model object (need to drop NA)
```

