---
title: "Feb 26 Notes"
author: "Claire Brandes"
date: "`r Sys.Date()`"
output: 
  prettydoc::html_pretty:
    theme: leonids
    toc: yes
---

#### Descriptive analysis example

```{r echo=FALSE}
library(readr)
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/tbs-2006-2008-ranges.csv"
d <- read_csv(f, col_names = TRUE)
```


```{r}
k95_males <- subset(d, sex.code == 2, kernel95)
k95_females <- subset(d, sex.code == 1, kernel95)
```


Determine mean, SD, and SE in "kernel95" home range size for each sex. 

Means:

```{r}
mean_M <- mean(k95_males$kernel95)
mean_F <- mean(k95_females$kernel95)

mean_M
mean_F
```

Standard Deviation: 

```{r}
sd_M <- sd(k95_males$kernel95)
sd_F <- sd(k95_females$kernel95)

sd_M
sd_F
```

Standard Error:

```{r}
SE_M <- sd_M/sqrt(9)
SE_F <- sd_F/sqrt(11)

SE_M
SE_F
```

Boxplot comparing "kernel95" by sex:

```{r}
library(ggplot2)
ggplot(data = d, aes(x = sex, y = kernel95)) + geom_boxplot(na.rm = TRUE) + xlab("Sex") + ylab("kernel95") + theme_classic()
```

For each sex, generate a bootstrap distribution, resampling data within each sex 10,000 times.

Females:

```{r}
n_boot <- 10000
boot_F <- vector(length = n_boot)
n_female <- length(k95_females$kernel95)

for (i in 1:n_boot) {
  boot_F[[i]] <- mean(sample(k95_females$kernel95, n_female, replace = TRUE))
}
```

Males:

```{r}
boot_M <- vector(length = n_boot)
n_male <- length(k95_males$kernel95)

for (i in 1:n_boot) {
  boot_M[[i]] <- mean(sample(k95_males$kernel95, n_male, replace = TRUE))
}
```

**Data Visualization:**

Female Histogram:

```{r}
hist(boot_F, breaks = 25, ylim = c(0, 1600), xlab = "Mean", ylab = "Frequency", main = "Female Bootstrap Distribution")
```

Male Histogram:

```{r}
hist(boot_M, breaks = 25, ylim = c(0, 1600), xlab = "Mean", ylab = "Frequency", main = "Male Bootstrap Distribution")
```

**Calculating Confidence Intervals:**

Males:

```{r}
lowerM <- quantile(boot_M, 0.025)
upperM <- quantile(boot_M, 0.975)
(ci_bootM <- c(lowerM, upperM)) # not sure if this is correct
```

Females:

```{r}
lowerF <- quantile(boot_F, 0.025)
upperF <- quantile(boot_F, 0.975)
(ci_bootF <- c(lowerF, upperF))
```

#### Inferential Analysis Example

Is there a significant difference in mean home range size of males and females?

What is H~0~? What is H~a~?

Two sample t-test - What is the test statistic?

```{r}
sp <- ((9-1) * sd(k95_males$kernel95)^2 + (11 - 1) * sd(k95_females$kernel95)^2) / (9 + 11 -2)
sp
```

```{r}
t_stat <- (mean(k95_males$kernel95) - mean)
((k95_females$kernel95) - 0) / sqrt(sp * 1/11 + 1) # not correct, need to fix 
```

```{r}
t.test(k95_females, k95_males, alternative = "two.side") # this shows correct t-statistic 
```

